MSc Thesis

### 2023-08-23 manuscript/main.tex
I think it is ready!

\listfiles
\documentclass[12pt, letterpaper, ]{report}
\usepackage[style=authoryear, maxnames=2, maxbibnames=99, citestyle=apa]{biblatex}
\addbibresource{merged_biblio2.bib}
\addbibresource{internet_stuff.bib}
%\addbibresource{induction.bib}
%\addbibresource{gwas_silicon.bib}
\usepackage{caption}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{url}
\usepackage{lineno}
\usepackage{tabularx}
\usepackage{gensymb}
\usepackage{longtable}
\usepackage{titlesec}
\usepackage{subcaption}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{array}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyfoot[R]{\thepage}
\fancypagestyle{plain}{%
    \renewcommand{\headrulewidth}{0pt}%
    \fancyhf{}%
    \fancyfoot[R]{\thepage}%
}

\newcommand{\mycomment}[1]{}
% Change chapter heading font size to 14pt
\titleformat{\chapter}{\Large\bfseries}{\thechapter.}{20pt}{\Large}

% Change section heading font size to 12pt
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{\large}

% Change subsection heading font size to 10pt
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{\normalsize}

\linenumbers
\graphicspath{./images/}
\geometry{
        letterpaper,
        top = 1.27cm,
        bottom = 1.27cm,
        left = 2.54cm,
        right = 2.54cm,
        }
\title{MSc Thesis}
\author{Isaac Peetoom Heida}
\date{December 2022}
\doublespace

\begin{document}
\pagenumbering{roman}
\input{title.tex}

\tableofcontents
\newpage
\listoftables
\newpage
\listoffigures
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

One of the most promising developments to arise over the past 30 years is the potential of silicon to improve crop plant performance. With benefits to multiple dimensions of crop performance, silicon may be a key tool to guard crop production against uncertain future growing conditions. Our ability to mobilize silicon-based crop improvement strategies depends on a thorough understanding of the proximate and ultimate causes of silicon accumulation, including both the ecological and genetic interactions that can trigger increased uptake. In this thesis, I extend recent advances in our understanding of silicon ecology in cereal crops, testing for the presence of rapid silicification in common Canadian crops, and use a  genome-wide association study to identify genetic markers associated with high silicon content. We failed to observe rapid silicification in the tested crops. This may could be a result of a soil environment incapable of supplying sufficient silicon resources for rapid accumulation, or could reflect the effect of herbivore identity on specific defensive outcomes. In our genetic analysis, we failed to find genetic markers with significant associations to silicon content, but did find a marker that had a significant correlation with manganese content. Future work in this field might develop a mapping population using crosses from high and low silicon genotypes in this study, yielding experimental designs with increased power and thus may find QTLs associated with silicon content. Applications silicon research into crop production techniques are still limited by crucial gaps in our understanding of how ecology and genetics control silicon uptake, but this work provides yet another stepping stone towards the broader adoption of silicon into agricultural systems.

@@

Silicon provides tremendous benefits to plant health, but is not yet widely utilized in agriculture despite it being highly abundance and naturally occurring in most soils. Silicon acts a defensive compound against insects, and also helps plants deal with stressful growing conditions, but is currently not widely applied in agriculture. Using cereal crops and a wild ancestor of wheat, I attempted to identify ecological and genetic factors that drive silicon content.  We failed to find genetic regions that exerted control over silicon content, but observed large variations in silicon content between the growing sites. We did not observe rapid silicon uptake in our tested cereal crops, complicating our understanding of when rapid uptake is achieved by plant species. Continuing to build upon this research will help move silicon crop technologies closer to feasibility, and could provide an environmentally friendly strategy to reduce pesticide use while maintaining yields.

@@

The research presented in this thesis is original and unpublished. Isaac Peetoom Heida and Dr. Juli Carrillo, with assistance from Dr. Jean-Thomas Cornelis and Dr. Gurcharn Singh Brar, conceptualized and developed the experiment presented in Chapter Two. Isaac Peetoom Heida, Dr. Juli Carrillo and Dr. Gurcharn Singh Brar, with input from Dr. Jean-Thomas Cornelis, conceptualized and developed the experiment presented in Chapter Three.
Isaac Peetoom Heida developed the question and methodology for Chapter Two. Dr. Aaron Beattie, Dr. Mazen Aljarrah, and Dr. Gurcharn Brar provided seeds for the experiment. Isaac Peetoom Heida designed and set up the experiment, processed and analysed the samples, and performed the statistical analysis. Dr. Shaun Barker and the Mineral Deposit Research Unit of the University of British Columbia provided facilities and expertise for the XRF analysis of the tissue samples. Dr. Simone Castellarin and Dr. Joana Pico Carbajo provided facilities and guidance in developing the protocol for phenolic analysis. Chelsea Gowton provided indispensable assistance with phenolic sample processing. 
For Chapter Three, Isaac Peetoom Heida led planting, plot setup, and maintenance, with assistance from Grace Wang, Vincent Fetterley, Sara Salad, Katherine Buchanan, Martina Clausen, Paul Fisher, and Matt Tsuruda. Isaac Peetoom Heida led the sample harvest, processing and analysis. Kelly Wang, Grace Wang, and Chelsea Gowton assisted with the sample harvesting. Dr. Daria Reshetniak, Paul Fisher, Lucas Friesen, Katie Pryer, Dr. Kinga Treder, Chelsea Gowten, Dennis Chiu, Carly MacGregor, and Grace Wang all provided invaluable assistance with sample preparation. 
I use the first person plural for chapters two and three as they are intended for submission to a peer-reviewed journal.

@@

This work would not have happened without the continuous support of my committee, the members of the PIEE lab, my friends and family, and the generosity of a host of funding agencies. Dr. Juli Carrillo has guided me to become a more well-rounded scientist and person, and has always believed in my abilities. I thank Dr. Jean-Thomas Cornelis for providing the silicon spark to kick this entire project off. Dr. Gurcharn Singh Brar lent his time and lab to me to support the work, and stoked my passion for grasses from a new angle. The members of the PIEE lab were more than generous with their time, and their continued interest in my project and general convivatilty provided the encouragement I needed to keep going with my research. In some ways this work is the culmination of a lifetime of support from my family, as they encouraged my curiosity about the natural world and sought out opportunities for me when I was too young (or too dumb) to seek them out for myself. Lastly, NSERC, CPBI, and the UBC provided generous scholarships and fellowships in support of my research and personal development.

\chapter*{Decication}
\addcontentsline{toc}{chapter}{Dedication}

To my Auntie Margaret, for showing me where the wetland was all those years ago.

@@

Silicon abounds in the earth’s crust, with various silicates, such as silicon dioxide (SiO\textsubscript{2}), comprising about 60\% of the crust by mass (\cite{holland_41_2014}). Nearly all terrestrial plants grow in soils containing silicon, and thus absorb nominal amounts through passive transport as the plant absorbs water (\cite{debona_silicons_2017}). Silicates occur in a variety of forms, and vary in their plant availability. Crystalline forms, such as quartz, are highly resistant to weathering, and are poor sources of plant-available silicon, while amorphous forms of SiO\textsubscript{2} are more available (\cite{fraysse_surface_2009}). As soils age, biogenic amorphous silicates, such as diatom testes and plant phytholiths, comprise an increasing share of the plant-available silicon (\cite{de_tombeur_plants_2020}). 
Plants interact with silicon on a variety of levels, mobilizing it from soil aggregates, transporting it into and throughout their bodies, and finally precipitating it out of their xylem into solid masses in the leaves and stems. Within the soil environment, silicon commonly exists in both crystalline (geologic) and amorphous (biogenic) forms (\cite{haynes_contemporary_2014}). Amorphous silicates can derive from previous plant material that has decayed in the soil, but also from marine and aquatic organisms such as diatoms. Globally, the silicon cycle involves silicates weathering out of terrestrial sediments, moving along water courses, and eventually being deposited in the sea, where it is incorporated into various plankton species, and eventually deposited in seafloor sediments. The continual exodus of silicon from terrestrial sediments over geologic timescales means as ecosystems age, plants become more and more central in the local silicon cycle, with much of the silicon in living plant tissue being recycled from previous plant material decaying in the soil (\cite{de_tombeur_plants_2020}).

@@

One of the most important advances in plant-silicon research was understanding the mechanisms through which plants acquire and transport silicon to specific location within the plant (\cite{coskun_controversies_2019}). Silicon’s most common form in soil solution is silicic acid (H2SiO4), which has a maximum solubility of around 2 mM (\cite{haynes_contemporary_2014}). In highly weathered soils with low nutrient availabilities, plants take a more active role in liberating nutrients, including silicon, for uptake. Organic acids and chelating agents, exuded from plant roots, pry tightly bound nutrients such as phosphorus and silicon from soil aggregates, increasing their availabilities for uptake into the root system (\cite{de_tombeur_silicon_2021-1}). Once dissolved in soil solution, silicon can be taken up by active or passive transport by the root system. While there is some evidence that small amounts of silicic acid can be transported during water uptake, this method of transport is insufficient to explain the larger amounts of silicon found in some plant families. Research in rice has identified four gene products that actively transport silicon into and through the plant body. Two of these (LSi1, LSi2) transport silicic acid from the soil into the roots, while the other two (LSi3, LSi6) act to unload silicic acid from the xylem into leaves and inflorescences (\cite{yamaji_orchestration_2015}). Orthologs of these proteins have been identified in other cereal crops, and additional analogous silicon transporter proteins have been discovered in the Cucurbitaceae (\cite{reynolds_silicon_2016}). Though not identified, there is a hypothesized fifth protein responsible for loading silicic acid into the xylem (\cite{farooq_silicon_2015}). Throughout the plant kingdom, clades vary significantly in their silicon content and expression of silicon transporter proteins (\cite{ma_chapter_2001}). The expression of these genes, or lack-there-of, can not only influence the total amount of silicon accumulated by the plant, but also its relative distribution, as knockout of LSi6 increases leaf silicon content while decreasing the silicon content of seed husks in rice (\cite{yamaji_transporter_2008}). Breeding for silicon content, improved exudate-based silicon mobilisation and silicon use-efficiency in crop plants may be crucial to improving crop performance under a changing climate (\cite{christian_breeding_2022,de_tombeur_silicon_2021}). However, we still have a relatively poor understanding surrounding the how genetics influence the silicon phenotype of a plant. Further investigations into how genotypic variation is reflected in the silicon content of plants can aid in the discovery of new genes involved in silicon accumulation, and may provide targets for silicon breeding programs.

@@

Plants deposit silicon in specialized silica cells, forming phytoliths (\cite{waterman_short-term_2021}). Silicon deposits taxa-specific morphologies, suggesting evolutionary pressure selecting for these structures to yield certain functions to the plant (\cite{piperno_phytoliths_2006}). In stems, these phytoliths are often long and narrow, oriented parallel with the shoot, and seem to increase structural rigidity(\cite{stromberg_functions_2016}). The use of silicon as a structural component represents a highly energetically efficient strategy, as silicon is ~10x cheaper on an energy unit basis to produce than lignin (\cite{stromberg_functions_2016}). Stem silicification has been investigated as it relates to lodging resistance in cereal crops, and silicon supplementation has been found to reduce the prevalence of lodging in rice and wheat (\cite{dorairaj_influence_2017,muszynska_mechanistic_2021}). In leaves, phytoliths are typically more stout, though they still increase the mechanical toughness of the leaf (\cite{simpson_still_2017}). This overall toughness, and abundance of phytoliths in leaves likely evolved to limit herbivore damage, rather than improve the growth characteristics as in stem phytoliths (\cite{stromberg_functions_2016}). Leaf silicon concentrations are negatively correlated with relative growth rate and survival in a number of invertebrate herbivores (\cite{juma_influence_2015,massey_silica_2006,mir_silicon_2019}). Phytoliths act by wearing down the mandibles of insect herbivores, thus reducing feeding efficiency (\cite{waterman_short-term_2021-1,mir_silicon_2019}), as well as reducing the digestive efficiency of the herbivore gut (\cite{hunt_novel_2008}). Interestingly, even in the absence of silicon, plants develop silica cells, and rapidly fill them when silicon becomes available (\cite{waterman_short-term_2021-1}). As phytoliths develop in the leaves, polymerisation of monosilicic acid is aided by interactions with proteins in the cell wall, which act as sites of nucleation (\cite{nawaz_phytolith_2019}). Silicon deposition in the leaves can happen on relatively short time scales, outpacing the accumulation of other defensive compounds such as phenolics (\cite{waterman_short-term_2021}). Thus, silicon-based defences in crop plants may be one of the first lines of active anti-herbivore defence, providing rapid and sensitive responses to herbivory.

In addition to the anti-herbivore benefits, silicon deposition seems to provide heightened disease resistance and stress tolerance. when supplemented with silicon, plants are generally more resistant to a wide spectrum of stressors, including soil salinity, soil metal toxicity, cold and heat stress, UV stress, water deficits, and phosphorus deficiencies (\cite{cooke_consistent_2016}). Silicon may create apoplastic barriers that seal and toughen the plant tissue, which might explain the range of plant traits that show a positive response to silicon (\cite{coskun_controversies_2019}). Under this apoplastic barrier hypothesis, silicon deposits reduce water loss and radiation/temperature damage, and also limit the spread of effector compounds, dampening the effects of fungal pathogen and herbivore excretions designed to interfere with plant defensive physiology (\cite{coskun_controversies_2019}). Empirical evidence shows that silicon is effective at limiting the growth of plant fungal pathogens (\cite{fauteux_silicon_2005}). Continuing to untangle the various mechanisms through which silicon delivers beneficial effects to plants is key to fully realizing the potential of silicon in sustainable agriculture.

@@

Much of today’s plant silicon work is indebted to the pioneering work of \textcite{jones_silica_1967}, who published a comprehensive study of silica in biotic systems, and the subsequent charting of silicon content across the plant kingdom by \textcite{takahashi_possibility_1990}). Epstein’s seminal \citeyear{epstein_silicon_1999} paper provided a comprehensive review of the state of knowledge in plant silicon, and has spurred a generation of researchers to extend the preliminary findings of the 20th century out across crop production systems and plant ecologies around the world (\cite{coskun_controversies_2019,hartley_ecology_2016,cooke_is_2011,christian_breeding_2022,de_tombeur_silicon_2021}). Silicon is best studied in the grass family (Poaceae) due to the comparatively high silicon content found in most members of the family (often over 1\% of dry weight), as well as the economic importance of domesticated species within the clade (\cite{reynolds_silicon_2016}). The domesticated grass species rice, maize, wheat, and barley alone account for one-third of the worlds’ total cultivated land area (\cite{faostat}). Silicon supplementation as an agricultural practice has been extensively studied in rice and sugar cane, as these crops tend to deplete soil silicon stocks, necessitating replenishment by application of silicon-rich amendments (\cite{haynes_contemporary_2014,meena_case_2014}). The critical depletion of silicon in these production systems has spurred much research activity targeting the biology of silicon, particularly in rice (\cite{deren_variable_1992-1,dai_genetic_2005,rodrigues_silicon_2004,ye_priming_2013}). In most temperate soils globally, Si is rarely truly limiting in soils, though certain forms of silicon are much more plant available than others (\cite{fraysse_surface_2009}). Thus, silicon research has been slower to develop in dryland temperate cereal crops, (but see \cite{rains_active_2006,neu_silicon_2017,ahmad_silicon_2016}). Great work can still be done to improve the manner and efficiency in which these temperate crops utilize the ample silicon available in their soils. Our ability to integrate silicon as a tool for production improvement in crop production is currently limited by a poor understanding of the genetic controls over silicon accumulation, as well as a limited understanding about the extent to which crops utilize silicon in pest-protection.

\chapter{Chapter 1: Testing rapid silicon accumulation in four cereal crops under real and artificial herbivory}

@@

To counter acute damage from herbivores, plants have developed a host of defensive strategies, ranging from changes to the body plan to the development of novel compounds to poison organisms that try to eat the plant (\cite{agrawal_plant_2006}). Due to the vastly different nature and ontogeny of various defensive strategies in plants, plant defences operate across a range of intensities and time scales, from short-term temporary activation, to long-lasting changes in the morphology of the plant (\cite{agrawal_plant_2006, karban_induced_1989}). In most scenarios, plants induce defences in response to an external cue, and build in intensity over time, with defensive hormone signals peaking approximately five hours after the initial induction event (\cite{schmelz_quantitative_2003}). Despite this rapid hormonal response, actual defensive phenotypes are slower to emerge, due to the higher costs of synthesizing defensive compounds, or developing new tissues like thorns and spines (\cite{karban_induced_1989}). Many defensive responses are also context dependent, where the identity of the damaging actor, the severity of damage, and a host of other factors interact to determine the final defensive response (\cite{waterman_simulated_2019}). The most effective defensive strategies should be those that can either prevent herbivory outright, or rapidly respond to limit damage. These same strategies are also the most promising for crop production, where pest damage represents both an economic and food security cost. Integrating better natural plant defences into crop production systems may be key to reducing the environmental impact of agriculture, but hinges upon a thorough understanding of plant defensive physiology.

>>>>

\documentclass[12pt, letterpaper]{report}
\usepackage[style=authoryear, maxnames=2, maxbibnames=99, citestyle=apa]{biblatex}
\addbibresource{merged_biblio2.bib}
\addbibresource{internet_stuff.bib}
\usepackage{caption}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{ulem}
\usepackage{url}
\usepackage{lineno}
\usepackage{tabularx}
\usepackage{gensymb}
\usepackage{longtable}
\usepackage{titlesec}
\usepackage{subcaption}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{array}
\usepackage{fancyhdr}
\usepackage{chemformula}
\usepackage[hidelinks]{hyperref}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[R]{\thepage}
\fancypagestyle{plain}{%
\fancyhf{}%
    \renewcommand{\headrulewidth}{0pt}%
    \fancyfoot[R]{\thepage}%
}

% Modify the fancy page style for chapters
\fancypagestyle{fancy}{%
    \fancyhf{}%
    \fancyfoot[R]{\thepage}%
}

\newcommand{\mycomment}[1]{}

\titleformat{\chapter}[block]{\Large\bfseries}{Chapter \thechapter.}{20pt}{\raggedright\Large}
\titlespacing{\chapter}{0pt}{5pt}{5pt}
% Adjust section spacing to prevent orphans
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{\large}
\titlespacing*{\section}{0pt}{\baselineskip}{0.5\baselineskip}

% Adjust subsection spacing to prevent orphans
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{\normalsize}
\titlespacing*{\subsection}{0pt}{\baselineskip}{0.5\baselineskip}

\graphicspath{./images/}
\geometry{
        letterpaper,
        top = 2.54cm,
        bottom = 2.54cm,
        left = 2.54cm,
        right = 2.54cm,
        }
\title{MSc Thesis}
\author{Isaac Peetoom Heida}
\date{December 2022}
\doublespace

\begin{document}
\pagenumbering{roman}
\setcounter{page}{2}
\input{title.tex}
\input{coverpage.tex}
% Rest of your document

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Over the past 30 years, the potential of silicon to improve crop plant performance has gained increasing recognition in the field of plant science. Silicon may be a key tool to guard crop production against uncertain future growing conditions by improving crop tolerance to abiotic and biotic stressors. In this thesis, I take recent advances in our understanding of silicon ecology and extend them into cereal crops, testing for the presence of rapid ($<24$ hour) silicification in common Canadian crops after defence induction events, including herbivory, and used a genome-wide association study to identify potential genetic markers associated with high silicon content. We found that overall, simulated herbivory, but not cricket herbivory, increased silicon leaf content. However, at the species level, we found no significant increases in silicon content in response to herbivory treatments. This may be due to a silicon-poor soil environment or could reflect the effect of herbivore identity on specific defensive outcomes. In our genetic analysis, we found no genetic markers with significant associations to silicon content but did find a marker that had a significant correlation with manganese content. Our limited results may reflect the polygenic nature of silicon content in plants, or strong environmental effects. Future studies could build upon our results by developing a mapping population using genotypes we identified as representing the extremes of silicon content. Applications of silicon research into crop production techniques are still limited by crucial gaps in our understanding of how ecology and genetics control silicon uptake, but this work provides yet another steppingstone towards the broader adoption of silicon into agricultural systems.

@@

Silicon provides tremendous benefits to plant health but is not yet widely utilized in agriculture despite it being highly abundant and naturally occurring in most soils. Silicon acts as a defensive compound against insects, and helps plants deal with stressful growing conditions, but is not currently widely applied in agriculture. Using cereal crops and a wild ancestor of wheat, I attempted to identify ecological and genetic factors that drive silicon content.  We failed to find genetic regions that exerted control over silicon content but observed large variations in silicon content between the growing sites. We did not observe short-term silicon uptake in our tested cereal crops, complicating our understanding of when short-term silicon accumulation is achieved by plant species. Continuing to build upon this research will help move silicon crop technologies closer to feasibility and could provide an environmentally friendly strategy to reduce pesticide use while maintaining yields.

@@

The research presented in this thesis is original and unpublished. Isaac Peetoom Heida and Dr. Juli Carrillo, with assistance from Dr. Jean-Thomas Cornelis and Dr. Gurcharn Singh Brar, conceptualized and developed the experiment presented in Chapter Two. Isaac Peetoom Heida, Dr. Juli Carrillo and Dr. Gurcharn Singh Brar, with input from Dr. Jean-Thomas Cornelis, conceptualized and developed the experiment presented in Chapter Three.
Isaac Peetoom Heida developed the question and methodology for Chapter Two. Dr. Aaron Beattie, Dr. Mazen Aljarrah, and Dr. Gurcharn Brar provided seeds for the experiment. Isaac Peetoom Heida designed and set up the experiment, processed and analysed the samples, and performed the statistical analysis. Dr. Shaun Barker and the Mineral Deposit Research Unit of the University of British Columbia provided facilities and expertise for the XRF analysis of the tissue samples. Dr. Simone Castellarin and Dr. Joana Pico Carbajo provided facilities and guidance in developing the protocol for phenolic analysis. Chelsea Gowton provided indispensable assistance with phenolic sample processing. 
For Chapter Three, Isaac Peetoom Heida led planting, plot setup, and maintenance, with assistance from Grace Wang, Vincent Fetterley, Sara Salad, Katherine Buchanan, Martina Clausen, Paul Fisher, and Matt Tsuruda. Isaac Peetoom Heida led the sample harvest, processing and analysis. Kelly Wang, Grace Wang, and Chelsea Gowton assisted with the sample harvesting. Dr. Daria Reshetniak, Paul Fisher, Lucas Friesen, Katie Pryer, Dr. Kinga Treder, Chelsea Gowton, Dennis Chiu, Carly MacGregor, and Grace Wang all provided invaluable assistance with sample preparation. 
I use the first-person plural for chapters two and three as they are intended for submission to a peer-reviewed journal.

\newpage
\tableofcontents
\newpage
\listoftables
\newpage
\listoffigures

@@

This work would not have happened without the continuous support of my committee, the members of the PIEE lab, my friends and family, and the generosity of a host of funding agencies. Dr. Juli Carrillo has guided me to become a more well-rounded scientist and person and has always believed in my abilities. I thank Dr. Jean-Thomas Cornelis for providing the silicon spark to kick this entire project off. Dr. Gurcharn Singh Brar lent his time and lab to me to support the work and stoked my passion for grasses from a new angle. The members of the PIEE lab were more than generous with their time, and their continued interest in my project and general conviviality provided the encouragement I needed to keep going with my research. In some ways this work is the culmination of a lifetime of support from my family, as they encouraged my curiosity about the natural world and sought out opportunities for me when I was too young (or too dumb) to seek them out for myself. Lastly, NSERC, CPBI, and the UBC provided generous scholarships and fellowships in support of my research and personal development.

\chapter*{}
\addcontentsline{toc}{chapter}{Dedication}

\noindent{\Large\textit{To my Auntie Margaret, for showing me where the wetland was all those years ago.}}

@@

Silicon abounds in the earth’s crust, with various silicates, such as silicon dioxide (\ch{SiO2}), comprising about 60\% of the crust by mass (\cite{holland_41_2014}). Nearly all terrestrial plants grow in soils containing silicon, and thus absorb at least nominal amounts through passive transport as the plant absorbs water (though many plants have active transport) (\cite{debona_silicons_2017}). Silicates occur in a variety of forms and vary in their plant availability. Crystalline forms, such as quartz, are highly resistant to weathering, and are poor sources of plant-available silicon, while amorphous forms of \ch{SiO2} are more available (\cite{fraysse_surface_2009}). 
Plants interact with silicon on a variety of levels, mobilizing it from soil aggregates, transporting it into and throughout their bodies, and finally precipitating it out of the xylem into solid masses in the leaves and stems. Within the soil environment, silicon commonly exists in both crystalline (geologic) and amorphous (biogenic) forms (Haynes 2014). Amorphous silicates can derive from previous plant material that has decayed in the soil, but also from marine and aquatic organisms such as diatoms. Young soils typically have high fractions of silicon-bearing materials, including clay minerals, though plant-available silicon peaks in middle-age soils (~120,000 years old) (\cite{de_tombeur_silicon_2020}). The continual leaching of silicon from terrestrial sediments over geologic timescales means as ecosystems age, plants become more and more central in the local silicon cycle, with much of the silicon in living plant tissue being recycled from previous plant material decaying in the soil (\cite{de_tombeur_plants_2020}).

@@

One of the most important advances in plant-silicon research was understanding the mechanisms through which plants acquire and transport silicon to specific locations within the plant (\cite{coskun_controversies_2019}). Silicon’s most common form in soil solution is silicic acid (\ch{H_4SiO_4}), which has a maximum solubility of around 2 mM (\cite{haynes_contemporary_2014}). In highly weathered soils with low nutrient availability, plants take a more active role in liberating nutrients, and other elements, including silicon, for uptake. Organic acids and chelating agents, exuded from plant roots, pry tightly bound nutrients and beneficial elements such as phosphorus and silicon from soil aggregates, increasing their availability for uptake into the root system (\cite{de_tombeur_silicon_2021-1}). Once dissolved in soil solution, silicon can be taken up by active or passive transport by the root system. While there is some evidence that small amounts of silicic acid can be transported during water uptake, this method of transport is insufficient to explain the larger amounts of silicon found in some plant families.

Research in rice has identified four gene products that actively transport silicon into and through the plant body. Two of these (LSi1, LSi2) transport silicic acid from the soil into the roots, while the other two (LSi3, LSi6) act to unload silicic acid from the xylem into leaves and inflorescences (\cite{yamaji_orchestration_2015}). Orthologs of these proteins have been identified in other cereal crops, and additional analogous silicon transporter proteins have been discovered in the Cucurbitaceae (\cite{reynolds_silicon_2016}). Though not identified, there is a hypothesized fifth protein responsible for loading silicic acid into the xylem (\cite{farooq_silicon_2015}). Throughout the plant kingdom, clades vary significantly in their silicon content and expression of silicon transporter proteins (\cite{ma_chapter_2001}). The expression of these genes, or lack-there-of, can not only influence the total amount of silicon accumulated by the plant, but also its relative distribution, as knockout of LSi6 increases leaf silicon content while decreasing the silicon content of seed husks in rice (\cite{yamaji_transporter_2008,yamaji_orchestration_2015}). Breeding for silicon content, improved exudate-based silicon mobilisation and silicon use-efficiency in crop plants may be crucial to improving crop performance under a changing climate (\cite{de_tombeur_silicon_2021,christian_breeding_2022}). However, we still have a relatively poor understanding surrounding how genetics influence the silicon phenotype of a plant. Further investigations into how genotypic variation is reflected in the silicon content of plants can aid in the discovery of new genes involved in silicon accumulation and may provide targets for silicon breeding programs.

@@

Plants deposit silicon in specialized silica cells, forming phytoliths (\cite{waterman_short-term_2021-1}). Silicon deposits show taxa-specific morphologies, suggesting evolutionary pressure selecting for these structures to yield certain functions to the plant (\cite{piperno_phytoliths_2006}). In stems, these phytoliths are often long and narrow, oriented parallel with the shoot, and seem to increase structural rigidity (\cite{stromberg_functions_2016}). The use of silicon as a structural component represents a highly energetically efficient strategy, as silicon is ~10x cheaper on an energy unit basis to produce than lignin (\cite{stromberg_functions_2016}). Stem silicification has been investigated as it relates to lodging resistance in cereal crops, and silicon supplementation has been found to reduce the prevalence of lodging in rice and wheat (\cite{dorairaj_influence_2017,muszynska_mechanistic_2021}). In leaves, phytoliths are typically stouter, though they still increase the mechanical toughness of the leaf (\cite{simpson_still_2017}). This overall toughness, and abundance of phytoliths in leaves is thought to have evolved to limit herbivore damage, rather than improve the growth characteristics as in stem phytoliths (\cite{stromberg_functions_2016}).

Leaf silicon concentrations are negatively correlated with relative growth rate and survival in several invertebrate herbivores (\cite{massey_silica_2006,juma_influence_2015,mir_silicon_2019}). Phytoliths act by wearing down the mandibles of insect herbivores, thus reducing feeding efficiency (\cite{mir_silicon_2019,waterman_short-term_2021-1}), as well as reducing the digestive efficiency of the herbivore gut (\cite{hunt_novel_2008}). Interestingly, even in the absence of silicon, plants develop silica cells, and rapidly fill them when silicon becomes available (\cite{waterman_short-term_2021-1}). As phytoliths develop in the leaves, polymerisation of monosilicic acid is aided by interactions with proteins in the cell wall, which act as sites of nucleation (\cite{nawaz_phytolith_2019}). Silicon deposition in the leaves can happen on relatively short time scales, outpacing the accumulation of other defensive compounds such as phenolics (\cite{waterman_short-term_2021}). Thus, silicon-based defences in crop plants may be one of the first lines of active anti-herbivore defence, providing rapid and sensitive responses to herbivory.

In addition to the anti-herbivore benefits, silicon deposition seems to provide heightened disease resistance and stress tolerance. When supplemented with silicon, plants are generally more resistant to a wide spectrum of stressors, including soil salinity, soil metal toxicity, cold and heat stress, UV stress, water deficits, and phosphorus deficiencies (\cite{cooke_consistent_2016}). Empirical evidence also shows that silicon is effective at limiting the growth of some fungal plant pathogens (\cite{fauteux_silicon_2005}). These wide-ranging positive effects may be due to silicon-based apoplastic barriers that seal and toughen the plant tissue (\cite{coskun_controversies_2019}). Under this apoplastic barrier hypothesis, silicon deposits reduce water loss and radiation/temperature damage, and limit the spread of effector compounds, dampening the effects of fungal pathogen and herbivore excretions designed to interfere with plant defensive physiology (\cite{coskun_controversies_2019}). Continuing to untangle the various mechanisms through which silicon delivers beneficial effects to plants is key to fully realizing the potential of silicon in sustainable agriculture.

@@

Much of today’s plant silicon work is indebted to the pioneering work of \textcite{jones_silica_1967}, who published a comprehensive study of silica in biotic systems, and the subsequent charting of silicon content across the plant kingdom by \textcite{takahashi_possibility_1990}. Epstein’s \citedate{epstein_silicon_1999} seminal paper provided a comprehensive review of the state of knowledge in plant silicon and has spurred a generation of researchers to extend the preliminary findings of the 20th century out across crop production systems and plant ecologies around the world (\cite{cooke_is_2011,hartley_ecology_2016,coskun_controversies_2019,de_tombeur_silicon_2021,christian_breeding_2022}). Silicon has been best studied in the grass family (Poaceae) due to the comparatively high silicon content found in most members of the family (often over 1\% of dry weight), as well as the economic importance of domesticated species within the clade (\cite{reynolds_silicon_2016}). The domesticated grass species rice, maize, wheat, and barley alone account for one-third of the worlds’ total cultivated land area (\cite{faostat}). Silicon supplementation as an agricultural practice has been extensively studied in rice and sugar cane, as these crops tend to be grown in soils naturally low in silicon, and rapidly deplete the remaining silicon stocks, necessitating replenishment by application of silicon-rich amendments (\cite{haynes_contemporary_2014,meena_case_2014}). The depletion of silicon in these production systems has spurred much research activity targeting the biology of silicon, particularly in rice (\cite{deren_variable_1992-1,dai_genetic_2005,rodrigues_silicon_2004,ye_priming_2013}). In most temperate soils globally, Si is rarely truly limiting in soils, though certain forms of silicon are much more plant available than others (\cite{fraysse_surface_2009}). Thus, silicon research has been slower to develop in dryland temperate cereal crops, (but see \cite{rains_active_2006,ahmad_silicon_2016,neu_silicon_2017}). Great work can still be done to improve the manner and efficiency in which these temperate crops utilize the ample silicon available in their soils. Our ability to integrate silicon as a tool for improving crop production is currently limited by a poor understanding of the genetic controls over silicon accumulation, as well as a limited understanding about the extent to which crops utilize silicon in pest-protection.

%\printbibliography

\chapter{Testing rapid silicon accumulation in four cereal crops under real and artificial herbivory}

@@

To counter acute damage from herbivores, plants have developed a host of defensive strategies, ranging from changes to the body plan to the development of novel compounds to poison organisms that try to eat the plant (\cite{agrawal_plant_2006}). Due to the vastly different nature and ontogeny of various defensive strategies in plants, plant defences operate across a range of intensities and time scales, from short-term temporary activation to long-lasting changes in the morphology of the plant (\cite{agrawal_plant_2006, karban_induced_1989}). In most scenarios, plants induce defences in response to an external cue, and build in intensity over time, with defensive hormone signals peaking approximately five hours after the initial induction event (\cite{schmelz_quantitative_2003}). Despite this rapid hormonal response, actual defensive phenotypes are slower to emerge, due to the higher costs of synthesizing defensive compounds, or developing new tissues like thorns and spines (\cite{karban_induced_1989}). Many defensive responses are also context dependent, where the identity of the damaging actor (e.g. species of insect), the severity of damage, and a host of other factors interact to determine the final defensive response (\cite{waterman_simulated_2019}). The most effective defensive strategies should be those that can either prevent herbivory outright, or rapidly respond to limit damage. These same strategies are also the most promising for crop production, where pest damage represents both an economic and food security cost. Integrating better natural plant defences into crop production systems could reduce some of the environmental impact of agriculture, but hinges upon a thorough understanding of plant defensive physiology.

One of the most promising avenues for new crop defence is the harnessing of silicon (\cite{reynolds_silicon_2016}). Silicon acts on multiple temporal and physiological scales, delivering broad spectrum resistance to pests, pathogens, and abiotic stressors (\cite{cooke_consistent_2016,coskun_controversies_2019}). Soluble silicon taken up from the soil is deposited predominantly in the leaf epidermis, where it forms solid granules that increase the toughness of the tissue, reducing herbivore digestive efficiency (\cite{cooke_is_2011}). Nearly all plants accumulate silicon over their lifespans, but there is also evidence of more acute silicon accumulation in response to herbivory (\cite{takahashi_possibility_1990}).
